---
title: "exploration"
author: "James Needell"
date: "2026-08-16"
output: pdf_document
---

```{r read_in}
# check layers
library(sf)
# data source URL: https://www.fs.usda.gov/rds/archive/catalog/RDS-2013-0009.7
gdb_path <- "RDS-2013-0009.7_Data_Format2_GDB/Data/FPA_FOD_20260615.gdb"
st_layers(gdb_path)
# read in dataset as fires. simple features(sf) and st_read() allows for gbd file to be recognized
fires <- st_read(gdb_path, layer = "Fires")
```



```{r tibbleizing}
library(tidyverse)
library(dplyr)
glimpse(fires)
Fires <- as.tibble(fires) # save as tibble
# view(Fires)
```

# What have I done so far?
1) set up GitHub repo
2) downloaded a data set
    - need to document source and citation information
3) read in data set with simple features library
4) saved as tibble and viewed it

# Goals
1) read AI assisted statistics for data scientists and work through it with real world data
2) read math for data science and work through it with python (using visual modeling to better understand concepts
or manipulate real world data, not sure yet...)
3) have this turn into a project stored on GitHub

# Next steps
1) read through statistics book and apply concepts to fires data set
    - probably first either create or import a data dictionary as a way to become familiar with the variables
2) simultaneously work through math for data science with python
    - this could be a different thing where I don't commit most of it to GitHub
    but maybe I'll just commit anyway for practice

Happy to have this set up and ready to start!

```{r data_dictionary}
# import csv provided with data
data_dict <- as.tibble(read_csv("RDS-2013-0009.7_Data_Format2_GDB/Data/_variable_descriptions.csv", show_col_types = FALSE))
# create a function to quickly look up variable description
desc_var <- function(col_name){
  data_dict %>%
    filter(Variable == col_name) %>%
    head(1) %>%
    pull(Description) %>%
    cat()
}
desc_var("Agency")
# view(data_dict)
```

“A basic step in exploring your data is getting a “typical value” for each feature (variable):
an estimate of where most of the data is located (i.e., its central tendency).” - AI assisted Stats

Estimates of Location
plan: choose a few variables and get an estimate of where the data is located or central tendency
and identify outliers
-mean
-weighted mean (ex: if we are taking data from sensors and one sensor is less accurate we may down weight it)
-median (robust)
-percentile (robust)
-weighted median (robust)
-trimmed mean (can be robust) usually preferable to ordinary mean


```{r estimates_of_location}
# what variables do I want to look at?
# fire size (categorical)
print(desc_var("FIRE_SIZE"))
# when the fire occured (start date)
print(desc_var("DISCOVERY_DATE"))
# where the fire occurred (categorical)
print(desc_var("STATE"))

# fire size or estimate of acres within the final perimeter of the fire
  # look by US first then could go state by state
fire_size <- Fires[["FIRE_SIZE"]] # index values in FIRE_SIZE col
mean_fs <- mean(fire_size)
median_fs <- median(fire_size)
trimmed_mean_fs <- mean(fire_size, 0.1)
# median is .72 and mean is 78 acres. This means there must be outliers
# trimmed mean at 10% is closer to median
median_fs # .72
mean_fs # 78
trimmed_mean_fs # 1.8188 at 10%
# in order to compute weighted mean and median but for what???
# could get population data > summarise by state, year, acres > join with population data >
# then the populated a state is, the more weight I can give to an acre

# claude code wrote in a csv with all state pop data from 2010-2025

  
# 2010-2020 data (drop 2020 here since the 2020s file has the more recent revision)
url_1 <- "https://www2.census.gov/programs-surveys/popest/datasets/2010-2020/national/totals/nst-est2020-alldata.csv"
pop_2010s <- read_csv(url_1) %>%
  filter(STATE != "00") %>%
  select(NAME, POPESTIMATE2010:POPESTIMATE2020) %>%
  pivot_longer(cols = starts_with("POPESTIMATE"),
               names_to = "year", names_prefix = "POPESTIMATE",
               values_to = "population") %>%
  filter(year != "2020")     # drop older-vintage 2020 estimate, keep the 2020s file's version instead

# 2020-2025 data
url_2 <- "https://www2.census.gov/programs-surveys/popest/datasets/2020-2025/state/totals/NST-EST2025-ALLDATA.csv"
pop_2020s <- read_csv(url_2) %>%
  filter(STATE != "00") %>%
  select(NAME, POPESTIMATE2020:POPESTIMATE2025) %>%
  pivot_longer(cols = starts_with("POPESTIMATE"),
               names_to = "year", names_prefix = "POPESTIMATE",
               values_to = "population")

# combine, map full state names to abbreviations, fix DC/PR manually, format
combined <- bind_rows(pop_2010s, pop_2020s) %>%
  mutate(
    state = state.abb[match(NAME, state.name)],
    state = case_when(
      NAME == "District of Columbia" ~ "DC",
      NAME == "Puerto Rico" ~ "PR",
      TRUE ~ state
    ),
    year = as.integer(year)
  ) %>%
  select(state, year, population) %>%
  distinct(state, year, .keep_all = TRUE) %>%   # safety net against any remaining duplicate year rows
  arrange(state, year)

write_csv(combined, "data_processed/state_population_2010_2025.csv")

state_pop_2010_2025 <- read_csv("data_processed/state_population_2010_2025.csv")

# which states had the most fires?
# which states had the most acres burned? 

```

```{r join with pop data 2010 to 2025}
# group by fire year and state then summarise fire size
Fires_2 <- 
  Fires %>%
  group_by(FIRE_YEAR, STATE) %>%
  summarise("Fire_Size" = sum(FIRE_SIZE), .groups = "drop") %>%
  filter(FIRE_YEAR >= 2010 & FIRE_YEAR <= 2025)
# tidy pop data to join
sp_2 <- 
  state_pop_2010_2025 %>%
  rename(
    "STATE" = state,
    "FIRE_YEAR" = year
  )
fires_pop <- left_join(Fires_2, sp_2, by = c("STATE", "FIRE_YEAR"))
# filter out NAs
fires_pop <- fires_pop %>%
  filter(!is.na(Fire_Size), !is.na(population))
# view(fires_pop)
# now I can practice weighted mean. I'll give higher pop states more weight
library(matrixStats)
# weighted mean of total acres burned per state per year weighted by population
weighted.mean(fires_pop[["Fire_Size"]], w = fires_pop[["population"]], na.rm = TRUE) # 236337.8
weightedMedian(fires_pop[["Fire_Size"]], w = fires_pop[["population"]], na.rm = TRUE) # 22347.05
# indicates there are outliers
```

# Estimates of Variability
- look at variability of fire sizes from year to year in colorado from 2010-2025
  - could compute median fire size each year
  - could look at number of fires per year
  - or could still look on national level...which states had most acres burned
      - maybe weighting with pop was wrong and could have done square acres per state
std deviation
IQR (interquartile range) difference between 25th and 75th percentile
MAD (median absolute deviation) more robust than standard deviation
    
```{r colorado}
CO_1 <- 
  Fires %>%
    filter(STATE == "CO") %>%
    group_by(FIRE_YEAR) %>%
    summarise(
      "acres_burned" = sum(FIRE_SIZE, na.rm = TRUE)
    )
sd(CO_1[["acres_burned"]])
IQR(CO_1[["acres_burned"]])
mad(CO_1[["acres_burned"]]) # scale of mad is adjusted to be on same scale as sd in R
# view(CO_1)


```

# Exploring Data Distribution
percentiles and quantiles
boxplots(based on quantiles)
frequency tables and histograms

```{r quant table and boxplots}
# get a quick quantile table
CO_1 %>%
  pull(acres_burned) %>%
  quantile(p = c(.05, 0.25, 0.5, 0.75, 0.95))
# if the year had 9920.88 total acres burned it was in the 5th percentile
# if 128704.22 acres burned the year was in the 75th percentile
# median is 46334.76 acres burned

# boxplot

# looking at fire size and acres from 1992 - 2024. can always come back and make this look good
# can clearly see that 2020 and 2002 had some of the biggest fires on record followed by 2018 then 2012
# ggplot(Fires %>%
#          filter(STATE == "CO"), aes(y = factor(FIRE_YEAR), x = FIRE_SIZE)) + # in order to bin FIRE_YEAR(an integer)
#   # I needed to factor it
#   geom_boxplot(staplewidth = 0.25) +
#   labs(
#     y = NULL,
#     x = "Fire Size (acres)"
#   )
# professionalized with Claude
ggplot(Fires %>% filter(STATE == "CO"),
       aes(y = factor(FIRE_YEAR), x = FIRE_SIZE)) +
  geom_boxplot(staplewidth = 0.25, fill = "#D9822B", color = "#4A4A4A", 
               outlier.alpha = 0.7, outlier.size = 1.5, outlier.color = "#8B3A1F") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Wildfire Size by Year in Colorado",
    subtitle = "Distribution of individual fire sizes, 1992–2024",
    x = "Fire Size (acres)",
    y = NULL,
    caption = "Source: USDA Forest Service FPA_FOD dataset"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "gray40", size = 11),
    plot.caption = element_text(color = "gray50", size = 8, hjust = 0),
    axis.text.y = element_text(size = 8, face = "bold"),
    panel.grid.minor = element_blank()
  )
# could do a facet wrap and look at western states
western_states <- c("WA", "OR", "CA", "NV", "ID", "MT", "WY", "UT", "CO", "AZ", "NM")
ggplot(Fires %>%
         filter(STATE %in% western_states),
       aes(y = factor(FIRE_YEAR), x = FIRE_SIZE)) +
  geom_boxplot(staplewidth = 0.25, fill = "#D9822B", color = "#4A4A4A", 
               outlier.alpha = 0.7, outlier.size = 1.5, outlier.color = "#8B3A1F") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Wildfire Size by Year",
    subtitle = "Distribution of individual fire sizes, 1992–2024",
    x = "Fire Size (acres)",
    y = NULL,
    caption = "Source: USDA Forest Service FPA_FOD dataset"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "gray40", size = 11),
    plot.caption = element_text(color = "gray50", size = 8, hjust = 0),
    axis.text.y = element_text(size = 7, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  facet_wrap(~ STATE, scales = "free")
#  a binned scatterplot may be better here but this gives a general idea on how big fires get in each state

# DOES LARGEST FIRE OF THE YEAR CORRELATE WITH TOTAL ACRES BURNED?
# HOW DO I LOOK AT LARGE FIRES AND HOW THEY PLAY INTO BAD FIRE YEARS?
# COULD LOOK AT TOTAL ACRE BAR CHARTS SIDE BY SIDE
# COULD ALSO CUT OUT 1990s MAYBE???
# CAN YOU HAVE SOME LARGE FIRES AND TOTAL ACRES THAT YEAR IS STILL LOW?
  
```

```{r frequency table- fire sizes in CO}
# frequency tables and histograms
# how common are big fires, small fires, medium etc??
# bin fire sizes
CO <- Fires %>%
  filter(STATE == "CO")
breaks <- seq(
   from = min(CO[["FIRE_SIZE"]]),
   to = max(CO[["FIRE_SIZE"]]),
   length = 30
 )
CO %>%
  mutate(
    FireSize = cut(FIRE_SIZE, breaks = breaks,  
      right = TRUE, include.lowest = TRUE)) %>%
  dplyr::count(FireSize, .drop = FALSE)

# bin the lowest bin
CO_2 <- CO %>%
  filter(FIRE_SIZE <= 7200)
breaks <- seq(
   from = min(CO_2[["FIRE_SIZE"]]),
   to = max(CO_2[["FIRE_SIZE"]]),
   length = 10000
 )
CO_2 %>%
  mutate(
    FireSize = cut(FIRE_SIZE, breaks = breaks,  
      right = TRUE, include.lowest = TRUE)) %>%
  dplyr::count(FireSize, .drop = FALSE) # .drop = FALSE includes the empty bins
# there are over 80,000 fires that are under .7 acres
# those should probably get removed for some analysis
# what is the lower bound of fires that should get removed?
# LOOK AT FIRE_SIZE_CLASS data is already binned. Interesting
```
Almost all fires recored are very small. There is also FIRE_SIZE_CLASS
variable which is a manual binning of fire sizes

## Histograms
- Empty bins are included in the graph.
- Bins are of equal width.
- The number of bins (or, equivalently, bin size) is up to the user.
- Bars are contiguous—no empty space shows between bars, unless there is an empty bin.

```{r histograms and density plots}
# histogram of fire size class counts
# ggplot(CO %>%
#          filter(FIRE_SIZE_CLASS != "A" & FIRE_SIZE_CLASS != "B" & FIRE_SIZE_CLASS != "C" & FIRE_SIZE_CLASS != "D", FIRE_SIZE_CLASS != "G"), aes(x = FIRE_SIZE)) +
#   geom_histogram(bins = 10)
# WON'T WORK DUE TO OUTLIERS so I cut off some classes. not telling me much here
# except 1) this is from 1992 - 2024 so 2) you can see that large fires are relatively rare

# I want to look at day of the year that fires start > can do hist and density of this 
# instead. makes much more sense. could even facet years etc. DISCOVERY_DOY
# could break days to 12, or could put month col in using case_when and grepl
# can also look at total fires across the years, then could facet western states
# would be a better line graph
ggplot(CO, aes(x = DISCOVERY_DOY)) +
  geom_histogram(fill = "darkorange3", bins = 50) +
  scale_x_continuous(
    breaks = c(0, 50, 100, 150, 200, 250, 300, 350)
  ) +
  scale_y_continuous(
    breaks = c(1000, 2000, 3000, 4000, 5000)
  ) +
  labs(
    x = "Discovery Day of Year",
    y = "Total Fires",
    title = "Distribution of Total Fires in Colorado each Day of the Year",
    subtitle = "Combined Data from 1992 to 2024",
    caption = "Source: USDA Forest Service FPA_FOD dataset"
    ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "gray40", size = 11),
    plot.caption = element_text(color = "gray50", size = 8, hjust = 0),
    panel.grid.minor = element_blank()
    )


# density plot which corresponds to plotting the histogram as a proportion rather than count
ggplot(CO, aes(x = DISCOVERY_DOY)) +
  geom_histogram(aes(y = after_stat(density)), fill = "darkorange3", bins = 50) + # tranform y values into density
  geom_density() + # layer density geom on top
  scale_x_continuous(
    breaks = c("Jan" = 0, "Feb" = 30, "Mar" = 60, "Apr" = 90, "May" = 120, "Jun" = 150, "Jul" = 180, "Aug" = 210, "Sep" = 240, "Oct" = 270, "Nov" = 300, "Dec" = 330) # obviously not best practice but quick
  ) +
  labs(
    x = "Fire Discovery Date",
    y = "Fire Density",
    title = "Density Distribution of Total Wildfires in Colorado each Day of the Year",
    subtitle = "Combined Data from 1992 to 2024",
    caption = "Source: USDA Forest Service FPA_FOD dataset"
    ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 8)),
    plot.subtitle = element_text(color = "gray40", size = 11, margin = margin(b = 12)),
    plot.caption = element_text(color = "gray50", size = 8, hjust = 0, margin = margin(t = 10)),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    panel.grid.minor = element_blank()
    )
ggsave("outputs/CO_daily_density_dist.png", width = 8, height = 6, dpi = 300)
# total area under the density curve = 1 so you can estimate probabilities 
# by calculating area under the curve between 2 points on the x axis
```
# Exploring Binary and Categorical Data
mode is most commonly occuring category or value in a dataset

```{r bar chart}
# could do year FIRE_SIZE_CLASS
# or could look at general causes of fires in CO
# lookup table for fire size class
fire_size_lookup <- c(
  A = "0-0.25 acres",
  B = "0.26-9.9 acres",
  C = "10.0-99.9 acres",
  D = "100-299 acres",
  E = "300-999 acres",
  F = "1000-4999 acres",
  G = "5000+ acres"
)
library(forcats)
CO %>%
  select(NWCG_GENERAL_CAUSE, FIRE_SIZE_CLASS) %>%
  filter(FIRE_SIZE_CLASS != "A") %>%
  ggplot(aes(y = fct_rev(fct_infreq(NWCG_GENERAL_CAUSE)))) + # factoring cat var then ordering
  geom_bar(aes(fill = FIRE_SIZE_CLASS)) +
  scale_fill_discrete(
    name = "Fire Size",
    labels = fire_size_lookup # change legend labels using lookup table created earlier
    ) +
  labs(
    y = "Cause",
    x = "Count",
    title = "Causes of Fires in Colorado from 1992-2024",
    caption = "Source: USDA Forest Service FPA_FOD dataset"
  )
# can see most fires are natural (mode is natural for cause) and most are small(mode is A for size)
# small fires do make the data hard to visualize
# could look at cause by year. see if trends are changing etc

CO %>%
  select(NWCG_GENERAL_CAUSE, FIRE_SIZE_CLASS, FIRE_YEAR) %>%
  filter(FIRE_YEAR >= 2015) %>%
  ggplot(aes(y = fct_rev(fct_infreq(NWCG_GENERAL_CAUSE)))) + # factoring cat var then ordering
  geom_bar(aes(fill = FIRE_SIZE_CLASS)) +
  scale_fill_discrete(
    name = "Fire Size",
    labels = fire_size_lookup # change legend labels using lookup table created earlier
    ) +
  labs(
    y = "Cause",
    x = "Count",
    title = "Causes of Fires in Colorado from 1992-2024",
    caption = "Source: USDA Forest Service FPA_FOD dataset"
  ) +
  facet_wrap(~FIRE_YEAR)
# should I look ay cause vs total fire size?
# look at large fires
CO %>%
  select(NWCG_GENERAL_CAUSE, FIRE_SIZE_CLASS, FIRE_YEAR) %>%
  filter(FIRE_YEAR >= 2015, FIRE_SIZE_CLASS == "F" | FIRE_SIZE_CLASS == "G") %>%
  ggplot(aes(y = fct_rev(fct_infreq(NWCG_GENERAL_CAUSE)))) + # factoring cat var then ordering
  geom_bar(aes(fill = FIRE_SIZE_CLASS)) +
  scale_fill_discrete(
    name = "Fire Size",
    labels = fire_size_lookup # change legend labels using lookup table created earlier
    ) +
  labs(
    y = "Cause",
    x = "Count",
    title = "Causes of Large Wildfires in Colorado",
    subtitle = "Includes Wildfires from 1992-1994 of Class F and G",
    caption = "Source: USDA Forest Service FPA_FOD dataset"
  ) +
   theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 8)),
    plot.subtitle = element_text(color = "gray40", size = 11),
    plot.caption = element_text(color = "gray50", size = 8, hjust = 0, margin = margin(t = 10)),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
    )
  
ggsave("outputs/CO_wildfire_causes_FG.png", width = 8, height = 6, dpi = 300)

```
# Expected Value
really a form of weighted mean. It adds the ideas of future expectations and 
probability weights often based on subjective judgement. sum of values * their
probability of occurence

# Probability
the probability that an event will happen is the proportion of times it will occur if the situation could be repeated over and over, countless times

# Correlation Coefficient
gives an estimate of the correlation of two variables that always lies on the same scale
-1 perfect neg correlation, +1 perfect pos correlation, 0 no correlation
correlation table displays relationship between multiple variables but variables can still have a non-linear
relationship and have near 0 correlation coefficient

A dot B / norm A * norm B
really just calculting angle between 2 vectors
if correlation coefficient is 1 or -1 then the 2 vectors are linearly dependent

```{r correlation table}
library("corrplot")
# fire_size_class with fire size length(which I have to create)

CO_numeric <- CO %>%
  mutate(fire_duration = as.double((((CONT_DATE - DISCOVERY_DATE) / 60) / 60) / 24)) %>%
  select(fire_duration, FIRE_SIZE, FIRE_YEAR) %>% # select numeric variables to compare correlation
  drop_na()

cor_matrix <- cor(CO_numeric, use = "complete.obs", method = "spearman") # spearman based on rank of data so estimate is robust to outliers
cor_matrix

corrplot(cor_matrix, method = "color")

# could look at fire counts for western states and correlate those
# how to do fire counts
# each row is a fire?
CO_3 <- CO %>%
  group_by(FIRE_YEAR) %>%
  summarise(
    count = n(),
    total_size = sum(FIRE_SIZE, na.rm = TRUE)
    )
cor(CO_3)
corrplot(cor(CO_3), method = "color")
```
1) positive correlation between fire size and fire duration (.365) which is expected. Larger fires should take longer to put out.
2) negative correlation bewteen fire year and fire duration. this makes sense if fire fighting technology has improved over time.
3) negative correlation between fire size and fire year. again makes sense if fire fighting tech has improved over time. better detection, better containment methods. more efficient etc.
4) Fire count has a positive correlation over time and this may indicate that total number of fires per year have increased since 1992.
5) total size has a positive correlation over time which may indicate that while fires have overall been smaller total area burned has been greater and this maps with count increasing over time




```{r scatter plots and more cor plots}
r_value <- cor(CO_3$FIRE_YEAR, CO_3$count)
ggplot(CO_3, aes(x = FIRE_YEAR, y = count)) +
  geom_point(color = "darkorange3") +
  geom_smooth(color = "darkgrey") +
  scale_x_continuous(
    breaks = c(1992, 1996, 2000, 2004, 2008, 2012, 2016, 2020, 2024)
  ) +
  labs(
    x = NULL,
    y = "Total Fire Count",
    title = "Total Fire Count in Colorado from 1992 to 2024",
    subtitle = paste0("r = ", round(r_value, 3)),
    caption = "Source: USDA Forest Service FPA_FOD dataset"
    ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 8)),
    plot.caption = element_text(color = "gray50", size = 8, hjust = 0, margin = margin(t = 10)),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    panel.grid.minor = element_blank()
    )

# now I could facet western states here. easy but computing R for each is trickier

western_states <- c("WA", "OR", "CA", "NV", "ID", "MT", "WY", "UT", "CO", "AZ", "NM")

fires_by_state_year <- Fires %>%
  filter(STATE %in% western_states) %>%
  group_by(STATE, FIRE_YEAR) %>%
  summarise(count = n(), .groups = "drop")

ggplot(fires_by_state_year, aes(x = FIRE_YEAR, y = count)) +
  geom_point(color = "darkorange3") +
  geom_smooth(color = "darkgrey") +
  facet_wrap(~ STATE, scales = "free_y") + # allows for different scales
  scale_x_continuous(
    breaks = c(1992, 2000, 2008, 2016, 2024)
  ) +
  labs(
    x = NULL,
    y = "Total Fire Count",
    title = "Total Fire Count by Year, Western States",
    caption = "Source: USDA Forest Service FPA_FOD dataset"
    ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 8)),
    plot.caption = element_text(color = "gray50", size = 8, hjust = 0, margin = margin(t = 10)),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    panel.grid.minor = element_blank()
    )
ggsave("outputs/wildfire_total_year_state.png", width = 8, height = 6, dpi = 300)

# now inspect correlation matrix and corrplot between states

fires_wide <- fires_by_state_year %>%
  select(STATE, FIRE_YEAR, count) %>%
  pivot_wider(names_from = STATE, values_from = count)

state_cor_matrix <- fires_wide %>%
  select(-FIRE_YEAR) %>%
  cor(use = "complete.obs")
state_cor_matrix

corrplot(state_cor_matrix, method = "color", type = "upper",
         addCoef.col = "black", tl.col = "black", diag = FALSE)

png("outputs/state_correlation_matrix.png", width = 8, height = 8, units = "in", res = 300)
corrplot(state_cor_matrix, method = "color", type = "upper",
         addCoef.col = "black", tl.col = "black", diag = FALSE)
dev.off()
```
# Multivariant Analysis
appropriate type of bivariant or multivariant analysis depends on nature of data:
numeric or categorical
Hexagonal Binning and Contours (numeric vs numeric, use when there are many datapoints when scatter plot is too dense)
Contingency table
  - table of counts by category

```{r hexplot}
CO_4 <- CO %>%
  mutate(fire_duration = as.double((((CONT_DATE - DISCOVERY_DATE) / 60) / 60) / 24))
ggplot(CO_4 %>%
         filter(FIRE_SIZE_CLASS != "A", FIRE_SIZE_CLASS != "B", FIRE_SIZE_CLASS != "C", fire_duration >= 2, FIRE_YEAR >= 2010), aes(x = fire_duration, y = FIRE_SIZE)) +
 geom_hex(color = "white") +
  scale_fill_gradient(low = "blue", high = "black") +
  facet_wrap(~FIRE_YEAR)


  
```

```{r contingency table}
# comparing 2 categorical variables
# cause and class
# library(gmodels)
# ct <- CrossTable(Fires$FIRE_SIZE_CLASS, Fires$NWCG_GENERAL_CAUSE,
#                  prop.c = FALSE, prop.chisq = FALSE, prop.t = FALSE)
# contingency table
cont_tibble <- Fires %>%
  dplyr::count(FIRE_SIZE_CLASS, NWCG_GENERAL_CAUSE) %>%
  group_by(FIRE_SIZE_CLASS) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  ungroup() %>%
  mutate(cell = paste0(n, " (", pct, "%)")) %>%
  select(-n, -pct) %>%
  pivot_wider(names_from = NWCG_GENERAL_CAUSE, values_from = cell, values_fill = "0 (0%)")

write_csv(cont_tibble, "outputs/class_cause_contingency.csv")
# for cat vs numeriacal look at duration vs agency



```

```{r more boxplots}
Fires_3 <- Fires %>%
  filter(STATE %in% western_states, !is.na(CONT_DATE), FIRE_YEAR == 2020) %>% # for many small fires cont date not input and have to focus on one year because fire fighting methods have improved over time
  mutate(STATE = factor(STATE)) %>%
  mutate(fire_duration = as.double((((CONT_DATE - DISCOVERY_DATE) / 60) / 60) / 24))
# Fires_3
ggplot(Fires_3, aes(x = STATE, y = fire_duration)) + 
  geom_boxplot() + 
  facet_wrap(~FIRE_SIZE_CLASS, labeller = as_labeller(fire_size_lookup)) 
# looking at these graphs I need to see counts to better understand data
# vibe coding with claude to get n on x axis
counts_per_panel <- Fires_3 %>%
  filter(!is.na(fire_duration)) %>%
  dplyr::count(STATE, FIRE_SIZE_CLASS)

Fires_3_labeled <- Fires_3 %>%
  left_join(counts_per_panel, by = c("STATE", "FIRE_SIZE_CLASS")) %>%
  mutate(state_label = paste0(STATE, "\n(n=", n, ")"))

ggplot(Fires_3_labeled, aes(x = state_label, y = fire_duration)) + 
  geom_boxplot() + 
  facet_wrap(~FIRE_SIZE_CLASS, labeller = as_labeller(fire_size_lookup), scales = "free_x") +
  labs(x = "State") +
  theme(axis.text.x = element_text(size = 7, angle = 45, hjust = 1))  

# now focus on 2 largest class sizes
ggplot(Fires_3_labeled %>%
         filter(FIRE_SIZE_CLASS == "F" | FIRE_SIZE_CLASS == "G"), aes(x = state_label, y = fire_duration)) + 
  geom_boxplot(fill = "#D9822B", color = "#4A4A4A", 
               outlier.alpha = 0.6, outlier.size = 1.2, outlier.color = "#8B3A1F", staplewidth = .25) +
  facet_wrap(~FIRE_SIZE_CLASS, labeller = as_labeller(fire_size_lookup), scales = "free_x") +
  labs(
    title = "Duration of the Largest Wildfires by Western US States",
    subtitle = "2020 — fire size classes F and G",
    x = "State",
    y = "Duration (days)",
    caption = "Source: USDA Forest Service FPA_FOD dataset"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 15, margin = margin(b = 4)),
    plot.subtitle = element_text(color = "gray40", size = 11, margin = margin(b = 12)),
    plot.caption = element_text(color = "gray50", size = 8, hjust = 0, margin = margin(t = 10)),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    axis.text.x = element_text(size = 7, angle = 45, hjust = 1),
    strip.text = element_text(size = 9, face = "bold"),
    panel.grid.minor = element_blank(),
    panel.spacing = unit(1, "lines")
  ) 
  ggsave("outputs/fire_duration_2020_FG.png", width = 8, height = 6, dpi = 300) # saves most recent file to outputs as png
# rain, wind, weather could influence this along with many other factors

# another option is a violin plot - can show nuances in distribution that aren't perceptable in boxplot
# density boxplot basicallly

```

Are fires getting bigger over time in colorado, western states?

```{r fire size over time}
ggplot(Fires %>% filter(FIRE_SIZE > 0, STATE == "CO"), aes(x = factor(FIRE_YEAR), y = FIRE_SIZE)) +
  geom_violin(fill = "#D9822B", color = "#4A4A4A") +
  scale_y_log10(breaks = c(0.1, 1, 10, 100, 1000, 10000, 100000), labels = scales::comma) +
  labs(x = "Year", y = "Fire Size (acres, log scale)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# violin not useful
# plotting summary statistics over time
Fires %>%
  filter(FIRE_SIZE > 0, STATE == "CO") %>%
  group_by(FIRE_YEAR) %>%
  summarise(
    median_size = median(FIRE_SIZE, na.rm = TRUE),
    mean_size = mean(FIRE_SIZE, na.rm = TRUE),
    p90 = quantile(FIRE_SIZE, 0.9, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = FIRE_YEAR)) +
  geom_line(aes(y = median_size, color = "Median")) +
  geom_line(aes(y = mean_size, color = "Mean")) +
  geom_line(aes(y = p90, color = "90th percentile")) +
  scale_y_log10(labels = scales::comma) +
  labs(x = "Year", y = "Fire Size (acres, log scale)", color = "Statistic")

# median is wonky. reporting or something. too in the weeds for right now
# perhaps other ways to answer this question

```
```{r questionable}
Fires %>%
  group_by(FIRE_YEAR) %>%
  summarise(
    n = dplyr::n(),
    pct_under_1_acre = mean(FIRE_SIZE < 1, na.rm = TRUE),
    pct_under_point25 = mean(FIRE_SIZE <= 0.25, na.rm = TRUE)
  )
# ????
```



# SUMMARY

Exploratory data analysis (EDA), pioneered by John Tukey, set a foundation for the field of data science.
The key idea of EDA is that the first and most important step in any project based on data is to look at the data.
By summarizing and visualizing the data, you can gain valuable intuition and  understanding of the project.

This chapter has reviewed concepts ranging from simple metrics, such as estimates of location and variability, to rich visual displays that explore the relationships between multiple variables, as in Figure 1-12.
The diverse set of tools and techniques being developed by the open source community, combined with the expressiveness of the R and Python languages, has created a plethora of ways to explore and analyze data.
Exploratory analysis should be a cornerstone of any data science project.

In this chapter, we demonstrated how LLMs can be used to improve or create data dictionaries as well as to help interpret and improve visualizations.


Excerpt From
AI-Assisted Statistics for Data Scientists
Peter Bruce, Andrew Bruce & Peter Gedeck
https://books.apple.com/us/book/ai-assisted-statistics-for-data-scientists/id6781040434
This material may be protected by copyright.
