---
title: "exploration"
author: "James Needell"
date: "2026-08-16"
output: pdf_document
---

```{r read_in}
# check layers
library(sf)
gdb_path <- "RDS-2013-0009.7_Data_Format2_GDB/Data/FPA_FOD_20260615.gdb"
st_layers(gdb_path)
# read in dataset as fires. simple features(sf) and st_read() allows for gbd file to be recognized
fires <- st_read(gdb_path, layer = "Fires")
```



```{r tibbleizing}
library(tidyverse)
library(dplyr)
glimpse(fires)
Fires <- as.tibble(fires) # save as tibble
# view(Fires)
```

# what have I done so far?
1) set up GitHub repo
2) downloaded a data set
    - need to document source and citation information
3) read in data set with simple features library
4) saved as tibble and viewed it

# goals
1) read AI assisted statistics for data scientists and work through it with real world data
2) read math for data science and work through it with python (using visual modeling to better understand concepts
or manipulate real world data, not sure yet...)
3) have this turn into a project stored on GitHub

# next steps
1) read through statistics book and apply concepts to fires data set
    - probably first either create or import a data dictionary as a way to become familiar with the variables
2) simultaneously work through math for data science with python
  - this could be a different thing where I don't commit most of it to GitHub
    but maybe I'll just commit anyway for practice

Happy to have this set up and ready to start!

